Mixing time estimation in reversible Markov chains from a single sample path

### 2015-06-03 conclusions.tex
Aryeh's changes
mostly grammatical/spelling; nontrivial ones indicated by \todoa

%!TEX root =  matrix-est.tex
Generalizing the construction of  \cref{thm:combined}, 
given any confidence interval $I_n=I_n(\gap,\pimin)$ for the spectral gap $\gap$ that itself depends 
on $\gap$ and $\pimin$, under general conditions,
plugging in the empirical bounds of \cref{alg:empest} into the expression defining $I_n$
gives a fully empirical confidence interval whose
width matches that of $I_n$ up to lower order terms.
In particular, this suggests that future work should focus on 
closing the gap between the lower and upper bounds on the accuracy
of point-estimation. Another interesting direction is to 
reduce the computation cost: The current cubic cost in the number of states
can be too high even when the number of states is moderate.

Perhaps more important, however, is to extend 
our results to large state space Markov chains:
In most practical applications the state space is continuous
or is exponentially large in some natural parameters.
As follows from our lower bounds, without further assumptions,
the problem of fully data dependent estimation of the mixing time
is intractable for information theoretical reasons.
Interesting directions for future work thus must consider Markov
chains with specific structure. Parametric classes of Markov chains,
including but not limited to Markov chains with factored transition kernels
with a few factors, are a promising candidate for such future investigations.
While we realize the limitations of our work, 
we think that the area of fully data dependent estimation of Markov chain 
parameters based on a single sample path is full of exciting, important
and challenging research questions and we hope that our work will
serve as a source of inspiration for others to consider these questions.
\if0
The algorithm from Section~\ref{sec:empirical} demonstrates the
feasibility of fully empirical finite-sample confidence bounds on
Markov chain mixing time.
This makes it possible to apply learning methods that require
empirical bounds to many non-iid settings that are commonplace in many
applications.
%One limitation of our techniques is that they are limited to Markov
%chains over finite state spaces.
We expect that future results will generalize and tighten our
analysis, possibly extend it to large (or infinite) state spaces by
adding extra assumptions.

>>>>

%!TEX root =  matrix-est.tex
The construction of \cref{thm:combined}
suggests a general procedure for converting
any confidence interval $I_n=I_n(\gap,\pimin)$ for the spectral gap $\gap$,
which
%itself depends 
%on
is expressed in terms of
$\gap$ and $\pimin$,
to a fully empirical confidence interval whose
width matches that of $I_n$ up to lower order terms:
Plug in the empirical bounds of \cref{alg:empest} into the expression defining $I_n$.
\todoa{reworded}
In particular, this suggests that future work should focus on 
closing the gap between the lower and upper bounds on the accuracy
of point-estimation. Another interesting direction is to 
reduce the computation cost: The current cubic cost in the number of states
can be too high even when the number of states is moderate.

Perhaps more important, however, is to extend 
our results to large state space Markov chains:
In most practical applications the state space is continuous
or is exponentially large in some natural parameters.
As follows from our lower bounds, without further assumptions,
the problem of fully data dependent estimation of the mixing time
is intractable for information theoretical reasons.
Interesting directions for future work thus must consider Markov
chains with specific structure. Parametric classes of Markov chains,
including but not limited to Markov chains with factored transition kernels
with a few factors, are a promising candidate for such future investigations.
The results presented here are a first step in the ambitious research agenda
outlined above, and we hope that they will
%pave the way
serve as a point of departure
for
further insights
in the area of fully empirical estimation of Markov chain 
parameters based on a single sample path.
%is full of exciting, important
%and challenging research questions and we hope that our work will
%serve as a source of inspiration for others to consider these questions.
\todoa{rewrote ending}
\if0
The algorithm from Section~\ref{sec:empirical} demonstrates the
feasibility of fully empirical finite-sample confidence bounds on
Markov chain mixing time.
This makes it possible to apply learning methods that require
empirical bounds to many non-iid settings that are commonplace in many
applications.
%One limitation of our techniques is that they are limited to Markov
%chains over finite state spaces.
We expect that future results will generalize and tighten our
analysis, possibly extend it to large (or infinite) state spaces by
adding extra assumptions.

### 2015-06-03 intro-nips.tex
New discussion + minor edits
log -> ln
Aryeh's changes
mostly grammatical/spelling; nontrivial ones indicated by \todoa

This work considers the construction of fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The \emph{mixing time} $\tmix$ of the Markov chain is the time that it
takes for the chain to (approximately) reach its stationary
distribution,
\begin{align}
\label{eq:mixtimedef}
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\end{align}
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that captures the
mixing time $\tmix$ with the pre-specified probability $1-\delta$.

This problem is motivated by the numerous scientific applications and
machine learning tasks in which one is interested in estimating the
mean $\vpi(f) = \sum_i \pi_i f(i)$ of a function $f$ of the states of
a Markov chain.
This is the realms of the celebrated Markov Chain Monte Carlo (MCMC)
paradigm~\cite{liu2001monte}, but the problem also arises in
performance prediction involving time-correlated data, as is common in
reinforcement learning~\cite{sutton98}.
Observable bounds on mixing times are useful in the design and
diagnostics of these methods; they yield effective approaches to
assessing the estimation quality, even when \emph{a priori} knowledge
of the mixing time or correlation structure is unavailable.

@@

Although these asymptotic results help us understand the limiting
behavior of the sample mean over a Markov chain, they say little about
the finite-time non-asymptotic behavior, which is often needed for the
prudent evaluation of a method or even its algorithmic design
\cite{
MCMCDiscussion93%
,DBLP:conf/valuetools/KontoyiannisLM06%
,BBL06%
,MniSzeAu08%
,MauPo09%
,LiLiWaSt11:KWIK%
,flegal2011implementing%
,Gyori-paulin15%
,SwaJoa15:LoggedBandit%
}. %BaHaVa10:activelearning,ShaVo08:conformalpred
To address this need, numerous works have developed Chernoff-type
bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
invaluable tools for non-asymptotic probabilistic
analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
These probability bounds are larger than corresponding bounds for
independent and identically distributed (iid) data due to the temporal
dependence; intuitively, for the Markov chain to yield a fresh draw
$X_{t'}$ that behaves as if it was independent of $X_t$, one must wait
$\Theta(\tmix)$ time steps.
Note that the bounds generally depend on distribution-specific
properties of the Markov chain (e.g., $\vP$, $\tmix$, $\gap$), which are often
unknown \emph{a priori} in practice.
Consequently, much effort has been put towards estimating these
unknown quantities, especially in the context of MCMC diagnostics, in
order to provide data-dependent assessments of estimation
accuracy~\cite[e.g.,][]{MCMCDiscussion93,GaSmi00:eigval,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
However, these approaches generally only provide asymptotic
guarantees, and hence fall short of our goal of empirical bounds that
are valid with any finite-length sample path.

@@

with additional access to the Markov chain.
For example, with access to a sampling oracle that generates

>>>>

This work
tackles the challenge of constructing
\todoa{was:
considers the construction of}
fully empirical bounds on the
mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
time-homogeneous Markov chain on a finite state space $[d] :=
\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
Under this assumption, the chain converges to its unique stationary
distribution $\vpi =
(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
\[
  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
  \quad \text{for each $i \in [d]$} .
\]
The \emph{mixing time} $\tmix$ of the Markov chain is the
number of time steps required
%time that it
%takes
for the chain to
be within a fixed threshold of
%(approximately) reach
its stationary
distribution:
\todoa{slightly changed wording}
\begin{align}
\label{eq:mixtimedef}
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]}
    \Abs{
      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
    }
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\end{align}
Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
set $A$ by $\vpi$, and the supremum is over all possible initial
distributions $\vq$.
The problem studied in this work is the construction of a non-trivial
confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
[0,\infty]$, based only on the observed sample path
$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that
succeeds with probability $1-\delta$
in trapping
the value of the
mixing time $\tmix$.
\todoa{changed wording}

This problem is motivated by the numerous scientific applications and
machine learning tasks in which
the quantity of interest is
%one is interested in estimating
the
mean $\vpi(f) = \sum_i \pi_i f(i)$
for some function $f$ of the states of
a Markov chain.
\todoa{changed wording}
This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
paradigm~\cite{liu2001monte}, but the problem also arises in
performance prediction involving time-correlated data, as is common in
reinforcement learning~\cite{sutton98}.
Observable bounds on mixing times are useful in the design and
diagnostics of these methods; they yield effective approaches to
assessing the estimation quality, even when \emph{a priori} knowledge
of the mixing time or correlation structure is unavailable.

@@

Although these asymptotic results help us understand the limiting
behavior of the sample mean over a Markov chain, they say little about
the finite-time non-asymptotic behavior, which is often needed for the
prudent evaluation of a method or even its algorithmic design
\cite{
MCMCDiscussion93%
,DBLP:conf/valuetools/KontoyiannisLM06%
,BBL06%
,MniSzeAu08%
,MauPo09%
,LiLiWaSt11:KWIK%
,flegal2011implementing%
,Gyori-paulin15%
,SwaJoa15:LoggedBandit%
}. %BaHaVa10:activelearning,ShaVo08:conformalpred
To address this need, numerous works have developed Chernoff-type
bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
valuable tools for non-asymptotic probabilistic
analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,
%meroi2012-jap,
paulin15}.
These probability bounds are larger than corresponding bounds for
independent and identically distributed (iid) data due to the temporal
dependence; intuitively, for the Markov chain to yield a fresh draw
$X_{t'}$ that behaves as if it was independent of $X_t$, one must wait
$\Theta(\tmix)$ time steps.
Note that the bounds generally depend on distribution-specific
properties of the Markov chain (e.g., $\vP$, $\tmix$, $\gap$), which are often
unknown \emph{a priori} in practice.
Consequently, much effort has been put towards estimating these
unknown quantities, especially in the context of MCMC diagnostics, in
order to provide data-dependent assessments of estimation
accuracy~\cite[e.g.,][]{MCMCDiscussion93,GaSmi00:eigval,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
However, these approaches generally only provide asymptotic
guarantees, and hence fall short of our goal of empirical bounds that
are valid with any finite-length sample path.

@@

%with additional access 
when allowed more flexible
to the Markov chain.
\todoa{slightly changed wording}
For example, given a sampling oracle that generates
